---
title: "Subprocesses in R"
author: "Lukasz A. Bartnik"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Subprocesses in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(subprocess)
library(knitr)

eval <- .Platform$OS.type != "windows"
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = eval)
```

## Introduction

Since R is not really a systems-programming language[^systemslanguage]
some facilities present in other such languages (e.g. C/C++, Python)
haven't been yet brought to R. One of such features is process
management which is understood here as the capability to create,
interact with and control the lifetime of child processes.

The R package `subprocess` aims at filling this gap by providing the few
basic operations to carry out the aforementioned tasks. The
`spawn_subprocess()` function starts a new child process and returns a
handle which can be later used in `process_read()` and `process_write()`
to send and receive data or in `process_wait()` or `process_terminate()`
to stop the such a process.

The R `subprocess` package has been designed after the exemplary
Python package which goes by the same. Its documentation can be found
[here](https://docs.python.org/3/library/subprocess.html) and numerous
examples of how it can be used can be found on the Web.

The R `subprocess` package has been implemented for both **Linux** and
**Windows** but since the author doesn't have access to a MacOSX machine
the extent to which it is compatible with Mac is unknown.


[^systemslanguage]: "By systems programming I mean writing code that
directly uses hardware resources, has serious resource constraints,
or closely interacts with code that does." Bjarne Stroustrup, "The
C++ Programming Language"


## Design and Implementation

The main concept in the package is the __handle__ which holds process
identifiers and an __external pointer__ object which in turn is a handle
to a low-level data structure holding various system-level parameters
of the running sub-process.

A child process, once started, runs until it exits on its own or until
its killed. Its current state as well as its exit status can be obtained
by dedicated API.

Communication with the child process can be carried out over the three
standard streams: the standard input, standard output and standard error
output. These streams are intercepted on the child process' side and
redirected into three anonymous pipes whose other ends are held by the
parent process and can be accessed through the process __handle__.

In **Linux** these are regular pipes created with the `pipe()` system call
and opened in the __non-blocking__ mode. All communication takes place
on request and follows the usual OS rules (e.g. the sub-process will
sleep if its output buffer gets filled).

In **Windows** these pipes are created with the `CreatePipe()` call and
opened in the __blocking__ mode. **Windows** does not support
__non-blocking__ (__overlapped__ in **Windows**-speak) mode for anonymous
pipes. For that reason each stream has an accompanying reader thread.
Reader threads are separated from R interpreter, do not exchange memory
with the R interpreter and will not break the single-thread assumption
under which R operates.


## Example: SSH connection

In this simple example we start a `ssh` client and connect to the local
ssh server. The connection is password-less, based on the public key.
Then, we list files in the remote  account. The example can be run in
both **Linux** and **Windows**, as long as a `ssh` __daemon__ is running
on `localhost` and a correct path to the `ssh` client is passed to
`spawn_process()`.


Open the connection:

```{r}
library(subprocess)

ssh_path <- '/usr/bin/ssh'
handle <- spawn_process(ssh_path, c('-T', 'test@localhost'))
process_read(handle, timeout = 1000)
process_read(handle, 'stderr')
```
The first number in the output is the value returned by `process_write`
which is the number of characters written to standard input of the
child process. The final `character(0)` is the output read from the
standard error stream.

Next, list files in the remote account:

```{r}
process_write(handle, 'ls\n')
process_read(handle, timeout = 1000)
process_read(handle, 'stderr')
```

Finally, we close the connection by exiting the remote __shell__:

```{r}
process_write(handle, 'exit\n')
process_read(handle, timeout = 1000)
process_read(handle, 'stderr')
```

The last thing is making sure that the child process is no longer alive:

```{r}
process_poll(handle)
process_return_code(handle)
```

## Advanced techniques

### Terminating a child process

If the child process needs to be terminated one can choose to:

  - send a command on the standard input with `process_write()`
  - send the termination signal, `SIGTERM` (**Linux**, **Windows**)
  - send the kill signal, `SIGKILL` (**Linux** only)

Assume the child R process is hung and there is no way to stop it
gracefully. Again, this example can be run on **Windows** if `"R"` is
replaced with `"R.exe"`[^termination].

`process_wait(handle, 1000)` will wait for 1 second (1000 milliseconds)
for the process to exit and then return `NA`. `process_terminate()` will
give `R` a chance to exit graceully, and then `process_kill()` will force
it to exit.


```{r}
R_binary <- file.path(R.home("bin"), "R")
handle <- spawn_process(R_binary, c('--slave', '-e', 'library(subprocess); .Call("C_signal", 15L, "ignore"); Sys.sleep(1000)'))

# process is hung
process_poll(handle, 1000)
process_return_code(handle)

# ask nicely to exit
process_terminate(handle)
process_poll(handle, 1000)
process_return_code(handle)

# forced exit
process_kill(handle)
process_poll(handle, 1000)
process_return_code(handle)
```

We see that the child process remains running until it receives the
`SIGKILL` signal[^signal]. The final return code (exit status) is the
number of the signal that caused the child process to exit[^status].

[^termination]: In **Windows**, `process_terminate()` is an alias for
`process_kill()`. They both lead to immediate termination of the child
process.

[^signal]: The `.Call("C_signal")` in our example is a call to a hidden
C function that `subprocess` provides mainly for the purposes of this
example.

[^status]: See the `waitpid()` manual page, e.g. [here](https://linux.die.net/man/2/waitpid). 


## Sending a signal to the child process

The last topic we want to cover here is sending an arbitrary[^windowssignals]
signal to the child process. **Linux** signals can be listed by looking
at the `signals` variable present in the package. It is constructed
automatically when the package is loaded and its value on **Linux** is
different than on **Windows**. In the example below we see the first
three elements of the **Linux** list of signals.

```{r}
length(signals)
signals[1:3]
```


`signals` is a `list` and its values are exported under their names in
the `subprocess` namespace.

```{r}
ls(pattern = 'SIG', envir = asNamespace('subprocess'))
```


Now we can create a new child process and send an arbitrary using its
handle.

```{r}
R_binary <- file.path(R.home("bin"), "R")
handle <- spawn_process(R_binary, '--slave')

process_send_signal(handle, SIGUSR1)
```


[^windowssignals]: The only signal currently supported in **Windows** is
`SIGTERM`. In the future versions of this package two additional signals
will be supported: `CTRL_C_EVENT` and `CTRL_BREAK_EVENT`.
