---
title: "Introduction to Sub-Processes in R"
author: "Lukasz A. Bartnik"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to Sub-Processes in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(subprocess)
library(knitr)

knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Introduction

Since R is not really a systems-programming language[^systemslanguage]
some facilities present in other such languages (e.g. C/C++, Python)
haven't been yet brought to R. One of such features is process
management which is understood here as the capability to create,
interact with and control the lifetime of child processes.

The R package `subprocess` aims at filling this gap by providing the few
basic operations to carry out the aforementioned tasks. The
`spawn_subprocess()` function starts a new child process and returns a
handle which can be later used in `process_read()` and `process_write()`
to send and receive data or in `process_wait()` or `process_terminate()`
to stop the such a process.

The R `subprocess` package has been designed after the exemplary
Python package which goes by the same. Its documentation can be found
[here](https://docs.python.org/3/library/subprocess.html) and numerous
examples of how it can be used can be found on the Web.

The R `subprocess` package has been verified to run on **Linux**,
**Windows** and **MacOS**.


[^systemslanguage]: "By systems programming I mean writing code that
directly uses hardware resources, has serious resource constraints,
or closely interacts with code that does." Bjarne Stroustrup, "The
C++ Programming Language"


## Design and Implementation

The main concept in the package is the __handle__ which holds process
identifiers and an __external pointer__ object which in turn is a handle
to a low-level data structure holding various system-level parameters
of the running sub-process.

A child process, once started, runs until it exits on its own or until
its killed. Its current state as well as its exit status can be obtained
by dedicated API.

Communication with the child process can be carried out over the three
standard streams: the standard input, standard output and standard error
output. These streams are intercepted on the child process' side and
redirected into three anonymous pipes whose other ends are held by the
parent process and can be accessed through the process __handle__.

In **Linux** these are regular pipes created with the `pipe()` system call
and opened in the __non-blocking__ mode. All communication takes place
on request and follows the usual OS rules (e.g. the sub-process will
sleep if its output buffer gets filled).

In **Windows** these pipes are created with the `CreatePipe()` call and
opened in the __blocking__ mode. **Windows** does not support
__non-blocking__ (__overlapped__ in **Windows**-speak) mode for anonymous
pipes. For that reason each stream has an accompanying reader thread.
Reader threads are separated from R interpreter, do not exchange memory
with the R interpreter and will not break the single-thread assumption
under which R operates.


## Introduction to Examples

Before we move on to examples, let's define a few helper functions
that abstract out the underlying operating system. We will use them
throughout this vignette.

```{r helpers}
is_windows <- function () (tolower(.Platform$OS.type) == "windows")

R_binary <- function () {
  R_exe <- ifelse (is_windows(), "R.exe", "R")
  return(file.path(R.home("bin"), R_exe))
}
```

Just for the record, vignette has been built in
`r ifelse(is_windows(), "Windows", "Linux")`.

```{r platform}
ifelse(is_windows(), "Windows", "Linux")
```

Now we can load the package and move on to the next section.

```{r}
library(subprocess)
```


## Example: controlling chlid R process

In this example we spawn a new R process, send a few commands to its
standard input and read the responses from its standard output. First,
let's spawn the child process (and give it a moment to complete the
start-up sequence[^syssleep]):

```{r new_child}
handle <- spawn_process(R_binary(), c('--no-save'))
Sys.sleep(1)
```

[^syssleep]: Depending on the system load, R can take a few seconds
to start and be ready for input. This is true also for other processes.
Thus, you will see `Sys.sleep()` following `spawn_process()` in almost
every example in this vignette.

Let's see the description of the child process:
```{r}
print(handle)
```

And now let's see what we can find it the child's output:
```{r read_from_child}
process_read(handle, PIPE_STDOUT, timeout = 1000)
process_read(handle, PIPE_STDERR)
```

The first number in the output is the value returned by `process_write`
which is the number of characters written to standard input of the
child process. The final `character(0)` is the output read from the
standard error stream.

Next, we create a new variable in child's session. Please notice the
new-line character at the end of the command. It triggers the child
process to process its input.

```{r new_n}
process_write(handle, 'n <- 10\n')
process_read(handle, PIPE_STDOUT, timeout = 1000)
process_read(handle, PIPE_STDERR)
```

Now it's time to use this variable in a function call:

```{r rnorn_n}
process_write(handle, 'rnorm(n)\n')
process_read(handle, PIPE_STDOUT, timeout = 1000)
process_read(handle, PIPE_STDERR)
```

Finally, we exit the child process:

```{r quit_child}
process_write(handle, 'q(save = "no")\n')
process_read(handle, PIPE_STDOUT, timeout = 1000)
process_read(handle, PIPE_STDERR)
```

The last thing is making sure that the child process is no longer alive:

```{r verify_child_exited}
process_state(handle)
process_return_code(handle)
```

Of course there is little value in running a child R process since there
are multiple other tools that let you do that, like `parallel`, `Rserve`
and `opencpu` to name just a few. However, it's quite easy to imagine
how running a remote shell in this manner enables new ways of
interacting with the environment. Consider running a local shell:

```{r spawn_shell}
shell_binary <- function () {
  ifelse (tolower(.Platform$OS.type) == "windows",
          "C:/Windows/System32/cmd.exe", "/bin/sh")
}

handle <- spawn_process(shell_binary())
print(handle)
```

Now we can interact with the shell sub-process. We send a request to
list the current directory, then give it a moment to process the command
and produce the output (and maybe finish its start-up, too). Finally,
we check its output streams.

```{r interact_with_shell}
process_write(handle, "ls\n")
Sys.sleep(1)
process_read(handle, PIPE_STDOUT)
process_read(handle, PIPE_STDERR)
```


## Advanced techniques

### Terminating a child process

If the child process needs to be terminated one can choose to:

  - send a command on the standard input with `process_write()`
  - send the termination signal, `SIGTERM` (**Linux**, **Windows**)
  - send the kill signal, `SIGKILL` (**Linux** only)

Assume the child R process is hung and there is no way to stop it
gracefully. `process_wait(handle, 1000)` waits for 1 second (1000
milliseconds) for the child process to exit. It then returns `NA` and
`process_terminate()` gives `R` a chance to exit graceully. Finally,
`process_kill()` forces it to exit.


```{r signal_child}
sub_command <- "library(subprocess);subprocess:::signal(15,'ignore');Sys.sleep(1000)"
handle <- spawn_process(R_binary(), c('--slave', '-e', sub_command))
Sys.sleep(1)

# process is hung
process_wait(handle, 1000)
process_state(handle)

# ask nicely to exit; will be ignored in Linux but not in Windows
process_terminate(handle)
process_wait(handle, 1000)
process_state(handle)

# forced exit; in Windows the same as the previous call to process_terminate()
process_kill(handle)
process_wait(handle, 1000)
process_state(handle)
```

We see that the child process remains running until it receives the
`SIGKILL` signal[^signal]. The final return code (exit status) is the
number of the signal that caused the child process to exit[^status].

[^termination]: In **Windows**, `process_terminate()` is an alias for
`process_kill()`. They both lead to immediate termination of the child
process.

[^signal]: The `.Call("C_signal")` in our example is a call to a hidden
C function that `subprocess` provides mainly for the purposes of this
example.

[^status]: See the `waitpid()` manual page, e.g. [here](https://linux.die.net/man/2/waitpid). 


### Sending a signal to the child process

The last topic we want to cover here is sending an arbitrary[^windowssignals]
signal to the child process. Signals can be listed by looking at the
`signals` variable present in the package. It is constructed
automatically when the package is loaded and its value on **Linux** is
different than on **Windows**. In the example below we see the first
three elements of the **Linux** list of signals.

```{r show_three_signals}
length(signals)
signals[1:3]
```


All possible signal identifiers are supported directly from the
`subprocess` package. Signals not supported on the current platform
are set to `NA` and the rest have their OS-specific numbers as their
values.

```{r}
ls(pattern = 'SIG', envir = asNamespace('subprocess'))
```


Now we can create a new child process and send an arbitrary using its
handle.

```{r eval=FALSE}
handle <- spawn_process(R_binary, '--slave')

process_send_signal(handle, SIGUSR1)
```


[^windowssignals]: The list of signals supported in **Windows** is much
shorter than the list of signals supported in **Linux** and contains the
following three signals: `SIGTERM`, `CTRL_C_EVENT` and `CTRL_BREAK_EVENT`.
